---
title: "02_prepare_cities_analysis"
author: "Olaf König"
format: html
editor: visual
---

# City-level analysis preparation and statistical computations

## Objective

Transform the base environmental dataset into analysis-ready data with local quartile classifications and aggregate statistics by socio-economic groups within each city.

## Source data

-   **Environmental dataset** : `data_input/data_edit/sep_cross_noise_no2_ndvi.gpkg` (households with environmental exposures)
-   **City population data** : `data_input/data_raw/stats_cities.csv` (for city selection and ranking)
-   **Municipal boundaries** : `data_input/data_raw/sf_polg_swisstopo.gpkg` (spatial references)

## Processing steps

1.  **City selection** : Define top 10 and top 6 cities by population + English translations
2.  **Data preparation** : NDVI logic inversion (green access → lack of green access)
3.  **Local classifications** : Quartile/quintile/decile rankings within each city for SES and environmental variables
4.  **Aggregate statistics** : Compute exposure rates by socio-economic groups and environmental thresholds
5.  **Data transformation** : Convert to long format for analysis and visualization

## Final outputs

-   `data_input/data_edit/cities_reclassified.gpkg` - Points with local classifications and labels
-   `df_stats_long` - Aggregate statistics in tidy format for visualizations

------------------------------------------------------------------------

```{r setup}

library(tidyverse)
library(tidylog)
library(sf)
```

# 1. City selection and base data preparation

## 1.1 Import datasets

```{r}
# Main environmental dataset
sf <- read_sf("data_input/data_edit/sep_cross_noise_no2_ndvi.gpkg") %>% 
  select(ssep3:geom) %>% 
  select(id, bfs_nummer, everything())

# Municipal boundaries and city populations
sf_municipalities <- read_sf("data_input/data_raw/sf_polg_swisstopo.gpkg")

df_cities_pop <- read_csv("data_input/data_raw/stats_cities.csv") %>% 
  select(bfs_nummer, nom_officiel, POP_2023)
```

## 1.2 Define city lists and translations

```{r}
# Top 10 cities by population
list_cities_10 <- df_cities_pop %>% 
  slice_max(POP_2023, n = 10) %>% 
  pull(bfs_nummer)

# Top 6 cities by population  
list_cities_6 <- df_cities_pop %>% 
  slice_max(POP_2023, n = 6) %>% 
  pull(bfs_nummer)

# English translations for international publication
cities_translation <- tibble(
  bfs_nummer = c(261, 6621, 2701, 5586, 351, 230, 1061, 3203, 5192, 371),
  name_en = c(
    "Zurich", "Geneva", "Basel", "Lausanne", "Bern",
    "Winterthur", "Lucerne", "St. Gallen", "Lugano", "Biel/Bienne"
  )
)
```

## 1.3 Data preparation and NDVI logic inversion

```{r}
# Filter to top 10 cities and add English names
sf <- sf %>% 
  filter(bfs_nummer %in% list_cities_10) %>% 
  left_join(cities_translation)

# Invert NDVI logic: from "green access" to "lack of green access"
# This ensures all indicators point in same direction (higher = worse)
sf <- sf %>%
  mutate(
    ndvi_200m = 1 - ndvi_200m,
    ndvi_500m = 1 - ndvi_500m
  )
```

# 2. Local classification functions

## 2.1 Quartile classification function

```{r}
prepare_groupings <- function(df) {
  df %>%
    mutate(
      # SES local classifications
      ssep3d_local_4  = ntile(ssep3, 4),
      ssep3d_local_5  = ntile(ssep3, 5), 
      ssep3d_local_10 = ntile(ssep3, 10),
      
      # Environmental variables local classifications
      noise_4  = ntile(noise, 4),
      noise_5  = ntile(noise, 5),
      noise_10 = ntile(noise, 10),
      
      no2_4    = ntile(no2, 4),
      no2_5    = ntile(no2, 5),
      no2_10   = ntile(no2, 10),
      
      ndvi_200m_4  = ntile(ndvi_200m, 4),
      ndvi_200m_5  = ntile(ndvi_200m, 5),
      ndvi_200m_10 = ntile(ndvi_200m, 10)
    )
}
```

## 2.2 Statistical summary function

```{r}
summarise_by_grouping <- function(sf_commune, grouping_var) {
  sf_commune %>%
    st_drop_geometry() %>%
    group_by(bfs_nummer, .data[[grouping_var]]) %>%
    summarise(
      n = n(),
      # Noise exposure rates
      pct_above_50_noise = mean(noise > 50) * 100,
      pct_above_60_noise = mean(noise > 60) * 100,
      # NO₂ exposure rates
      pct_above_15_no2   = mean(no2 > 15) * 100,
      pct_above_20_no2   = mean(no2 > 20) * 100,
      # Limited green access rates (inverted NDVI)
      pct_above_50_ndvi  = mean(ndvi_200m > 0.5) * 100,
      pct_above_60_ndvi  = mean(ndvi_200m > 0.6) * 100,
      .groups = "drop"
    )
}
```

# 3. City-level statistical computations

## 3.1 Apply local classifications by city

```{r}
# Create city-specific datasets with local classifications
list_sf_by_commune <- sf %>%
  filter(bfs_nummer %in% list_cities_10) %>%
  { split(., .$bfs_nummer) }

# Apply grouping function to each city
list_sf_by_commune_grouped <- map(list_sf_by_commune, prepare_groupings)
```

## 3.2 Compute exposure statistics by socio-economic groups

```{r}
# Quintiles (5 groups)
df_stats_quintiles <- map_dfr(
  list_sf_by_commune_grouped,
  summarise_by_grouping,
  grouping_var = "ssep3d_local_5"
)

# Quartiles (4 groups)
df_stats_quartiles <- map_dfr(
  list_sf_by_commune_grouped,
  summarise_by_grouping,
  grouping_var = "ssep3d_local_4"
)

# Deciles (10 groups)
df_stats_deciles <- map_dfr(
  list_sf_by_commune_grouped,
  summarise_by_grouping,
  grouping_var = "ssep3d_local_10"
)
```

## 3.3 Merge results across classification levels

```{r}
# Add classification type labels
df_stats_deciles   <- mutate(df_stats_deciles,   quantile_type = "Deciles")
df_stats_quintiles <- mutate(df_stats_quintiles, quantile_type = "Quintiles")
df_stats_quartiles <- mutate(df_stats_quartiles, quantile_type = "Quartiles")

# Standardize column names
df_stats_deciles <- df_stats_deciles %>%
  rename(group = ssep3d_local_10) %>%
  select(bfs_nummer, group, quantile_type, matches("^pct_above"))

df_stats_quintiles <- df_stats_quintiles %>%
  rename(group = ssep3d_local_5) %>%
  select(bfs_nummer, group, quantile_type, matches("^pct_above"))

df_stats_quartiles <- df_stats_quartiles %>%
  rename(group = ssep3d_local_4) %>%
  select(bfs_nummer, group, quantile_type, matches("^pct_above"))

# Combine all classification levels
df_stats_all <- bind_rows(df_stats_deciles, df_stats_quintiles, df_stats_quartiles)
```

# 4. Data transformation and export

## 4.1 Convert to long format for analysis

```{r}
df_stats_long <- df_stats_all %>%
  pivot_longer(
    cols = matches("^pct_above"),
    names_to = "indicator",
    values_to = "value"
  ) %>%
  mutate(
    # Clean indicator labels for publication
    indicator_label = recode(indicator,
      "pct_above_50_noise" = "Exposed to noise >50 dB",
      "pct_above_60_noise" = "Exposed to noise >60 dB",
      "pct_above_15_no2"   = "Exposed to NO₂ >15 μg/m³",
      "pct_above_20_no2"   = "Exposed to NO₂ >20 μg/m³",
      "pct_above_50_ndvi"  = "Limited green access (NDVI < 0.5)",
      "pct_above_60_ndvi"  = "Limited green access (NDVI < 0.6)"
    )
  ) %>% 
  left_join(cities_translation)
```

## 4.2 Add indicator families for visualization

```{r}
df_stats_long <- df_stats_long %>%
  mutate(
    indicator_family = case_when(
      grepl("noise", indicator) ~ "Bruit",
      grepl("no2", indicator) ~ "NO2",
      grepl("ndvi", indicator) ~ "NDVI"
    )
  )
```

## 4.3 Export classified dataset

```{r}
# Recombine all city datasets with classifications
sf_grouped <- bind_rows(list_sf_by_commune_grouped)

# Add classification labels for extreme combinations
sf_grouped <- sf_grouped %>%
  mutate(
    # SES-Noise combinations
    ssep_noise_label = case_when(
      ssep3d_local_4 == 1 & noise_4 == 4 ~ "Low SES / High Noise",
      ssep3d_local_4 == 1 & noise_4 == 1 ~ "Low SES / Low Noise", 
      ssep3d_local_4 == 4 & noise_4 == 4 ~ "High SES / High Noise",
      ssep3d_local_4 == 4 & noise_4 == 1 ~ "High SES / Low Noise",
      TRUE ~ "Other combination"
    ),
    # SES-NO₂ combinations
    ssep_no2_label = case_when(
      ssep3d_local_4 == 1 & no2_4 == 4 ~ "Low SES / High NO₂",
      ssep3d_local_4 == 1 & no2_4 == 1 ~ "Low SES / Low NO₂",
      ssep3d_local_4 == 4 & no2_4 == 4 ~ "High SES / High NO₂", 
      ssep3d_local_4 == 4 & no2_4 == 1 ~ "High SES / Low NO₂",
      TRUE ~ "Other combination"
    ),
    # SES-Green access combinations
    ssep_ndvi_label = case_when(
      ssep3d_local_4 == 1 & ndvi_200m_4 == 4 ~ "Low SES / Low Green",
      ssep3d_local_4 == 1 & ndvi_200m_4 == 1 ~ "Low SES / High Green",
      ssep3d_local_4 == 4 & ndvi_200m_4 == 4 ~ "High SES / Low Green",
      ssep3d_local_4 == 4 & ndvi_200m_4 == 1 ~ "High SES / High Green", 
      TRUE ~ "Other combination"
    )
  )

# Export final classified dataset
sf_grouped %>% 
  write_sf("data_input/data_edit/cities_reclassified.gpkg")

cat("✅ Export complete: cities_reclassified.gpkg\n")
cat("📊 Dataset ready for analysis and visualization\n")
```

## 4.4 Quick validation

```{r}
# Verify city coverage
cat("Cities included:", length(unique(sf_grouped$bfs_nummer)), "\n")
cat("Total households:", nrow(sf_grouped), "\n") 
cat("Statistical summary ready for", nrow(df_stats_long), "indicator-group combinations\n")
```
