---
title: "03_datawrapper_final_charts"
author: "Olaf König"
format: html
editor: visual
---

# Datawrapper charts for environmental inequality analysis

## Objective

Export data to the 9 specific Datawrapper charts used in the final article. Each chart shows different aspects of environmental inequalities across Swiss cities.

## Source data

-   `data_input/data_edit/cities_reclassified.gpkg` - Cities with local classifications
-   `df_stats_long` - Aggregate statistics from previous script
-   National dataset for comparison charts

## Charts exported

-   **National comparison**: `Qsh2p` (national quartiles)
-   **City aggregation**: `BnMJJ` (cities bar chart)\
-   **City disparities**: `iLHkd` (lollipop chart)
-   **City scatterplots**: `rffGE`, `8ZPgH`, `qRJ8l`, `MJu0Z`, `HfB0J` (5 specific cities/indicators)
-   **Situation map**: `4bAiG` (cities location map)

------------------------------------------------------------------------

```{r setup}
library(tidyverse)
library(sf)
library(DatawRappr)
```

```{r load-data}
# Load classified cities data
sf_grouped <- read_sf("data_input/data_edit/cities_reclassified.gpkg")

# Recreate df_stats_long if not in memory (should be from previous script)
if (!exists("df_stats_long")) {
  source("02_prepare_cities_analysis.qmd") # Or load from saved file
}
```

# 1. National comparison charts

## 1.1 National quartiles baseline (Qsh2p)

```{r}
# Calculate national quartiles for comparison
sf_national <- read_sf("data_input/data_edit/sep_cross_noise_no2_ndvi.gpkg") %>% 
  select(ssep3:geom) %>% 
  select(id, bfs_nummer, everything())

df_national <- sf_national %>% st_drop_geometry()

dw_national_quartiles <- df_national %>% 
  mutate(
    ssep3_global_quartiles = ntile(ssep3, 4)
  ) %>% 
  group_by(ssep3_global_quartiles) %>% 
  summarise(
    pct_above_50_noise = mean(noise > 50) * 100,
    pct_above_15_no2   = mean(no2 > 15) * 100,
    .groups = "drop"
  ) %>% 
  rename(
    "Households exposed to noise >50 dB, in %" = pct_above_50_noise,
    "Households exposed to NO₂ >15 μg/m³, in %" = pct_above_15_no2
  ) %>% 
  pivot_longer(
    cols = starts_with("Households"),
    names_to = "indicator",
    values_to = "value"
  ) %>% 
  mutate(
    ssep3_global_quartiles_label = case_when(
      ssep3_global_quartiles == 1 ~ "Lowest",
      ssep3_global_quartiles == 2 ~ "Lower-middle", 
      ssep3_global_quartiles == 3 ~ "Upper-middle",
      ssep3_global_quartiles == 4 ~ "Highest",
      TRUE ~ as.character(ssep3_global_quartiles)
    )
  ) %>% 
  arrange(desc(ssep3_global_quartiles))

# Export to Datawrapper
dw_national_quartiles %>% 
  dw_data_to_chart("Qsh2p")

cat("✅ National quartiles exported to Qsh2p\n")
```

# 2. City aggregation charts

## 2.1 Cities aggregated bar chart (BnMJJ)

```{r}
# Aggregate across all 10 cities
dw_cities_bar_charts <- sf_grouped %>% 
  st_drop_geometry() %>% 
  group_by(ssep3d_local_4) %>% 
  summarise(
    pct_above_50_noise = mean(noise > 50) * 100,
    pct_above_15_no2   = mean(no2 > 15) * 100,
    pct_above_50_ndvi  = mean(ndvi_200m > 0.5) * 100,
    .groups = "drop"
  ) %>% 
  rename(
    "Households exposed to noise >50 dB, in %" = pct_above_50_noise,
    "Households exposed to NO₂ >15 μg/m³, in %" = pct_above_15_no2,
    'Households with limited green access: less than 50% of "green" surfaces within 200m, in %' = pct_above_50_ndvi
  ) %>% 
  pivot_longer(
    cols = starts_with("Households"),
    names_to = "indicator",
    values_to = "value"
  ) %>% 
  mutate(
    ssep3d_local_4_label = case_when(
      ssep3d_local_4 == 1 ~ "Lowest",
      ssep3d_local_4 == 2 ~ "Lower-middle", 
      ssep3d_local_4 == 3 ~ "Upper-middle",
      ssep3d_local_4 == 4 ~ "Highest",
      TRUE ~ as.character(ssep3d_local_4)
    )
  ) %>%
  arrange(desc(ssep3d_local_4))



# Export to Datawrapper
dw_cities_bar_charts %>%
  dw_data_to_chart("BnMJJ")

cat("✅ Cities aggregated bars exported to BnMJJ\n")
```

## 2.2 City disparities lollipop chart (iLHkd)

```{r}
# Calculate differences between lowest and highest SES quartiles
dw_cities_lollipop <- df_stats_long %>% 
  filter(
    quantile_type == "Quartiles",
    group %in% c(1, 4),
    indicator_label %in% c(
      "Exposed to noise >50 dB",
      "Exposed to NO₂ >15 μg/m³",
      "Limited green access (NDVI < 0.5)"
    )
  ) %>% 
  select(bfs_nummer, name_en, group, indicator_label, value) %>% 
  mutate(
    indicator_label_en = case_when(
      indicator_label == "Exposed to noise >50 dB" ~ "Households exposed to noise >50 dB, in %",
      indicator_label == "Exposed to NO₂ >15 μg/m³" ~ "Households exposed to NO₂ >15 μg/m³, in %",
      indicator_label == "Limited green access (NDVI < 0.5)" ~ 'Households with limited green access: less than 50% of "green" surfaces within 200m, in %'
    )
  ) %>% 
  pivot_wider(
    names_from = "group",
    values_from = "value"
  ) %>% 
  mutate(
    difference = `1` - `4`
  ) %>% 
  rename(
    "Lowest SES" = `1`,
    "Highest SES" = `4`
  )



# Export to Datawrapper
dw_cities_lollipop %>%
  dw_data_to_chart("iLHkd")

cat("✅ City disparities lollipop exported to iLHkd\n")
```

# 3. City-specific scatterplots

## 3.1 St-Gallen NO₂ scatterplot (rffGE)

```{r}
sf_grouped %>%
  st_drop_geometry() %>%
  filter(bfs_nummer == 3203) %>%  # St-Gallen
  select(ssep3, no2, ssep_no2_label) %>%
  dw_data_to_chart("rffGE")


cat("✅ St-Gallen NO₂ scatterplot exported to rffGE\n")
```

## 3.2 Geneva NO₂ scatterplot (8ZPgH)

```{r}
sf_grouped %>% 
  st_drop_geometry() %>%
  filter(bfs_nummer == 6621) %>%  # Geneva
  select(ssep3, no2, ssep_no2_label) %>% 
  dw_data_to_chart("8ZPgH")

cat("✅ Geneva NO₂ scatterplot exported to 8ZPgH\n")
```

## 3.3 Geneva NDVI scatterplot (qRJ8l)

```{r}
sf_grouped %>% 
  st_drop_geometry() %>%
  filter(bfs_nummer == 6621) %>%  # Geneva
  select(ssep3, ndvi_200m, ssep_ndvi_label) %>% 
  mutate(ndvi_200m = ndvi_200m * 100) %>%  # Convert to percentage
  dw_data_to_chart("qRJ8l")

cat("✅ Geneva NDVI scatterplot exported to qRJ8l\n")
```

## 3.4 Lausanne noise scatterplot (MJu0Z)

```{r}
sf_grouped %>% 
  st_drop_geometry() %>%
  filter(bfs_nummer == 5586) %>%  # Lausanne
  select(ssep3, noise, ssep_noise_label) %>% 
  dw_data_to_chart("MJu0Z")

cat("✅ Lausanne noise scatterplot exported to MJu0Z\n")
```

## 3.5 Lausanne NDVI scatterplot (HfB0J)

```{r}
sf_grouped %>% 
  st_drop_geometry() %>%
  filter(bfs_nummer == 5586) %>%  # Lausanne
  select(ssep3, ndvi_200m, ssep_ndvi_label) %>% 
  mutate(ndvi_200m = ndvi_200m * 100) %>%  # Convert to percentage
  dw_data_to_chart("HfB0J")

cat("✅ Lausanne NDVI scatterplot exported to HfB0J\n")
```

# 4. Situation map data export

## 4.1 Cities situation map (4bAiG)

```{r}
# Recreate city centroids and population data for situation map
sf_municipalities <- read_sf("data_input/data_raw/sf_polg_swisstopo.gpkg")
df_cities_pop <- read_csv("data_input/data_raw/stats_cities.csv") %>% 
  select(bfs_nummer, nom_officiel, POP_2023)

list_cities_10 <- df_cities_pop %>% 
  slice_max(POP_2023, n = 10) %>% 
  pull(bfs_nummer)

sf_cities_10 <- sf_municipalities %>% 
  filter(bfs_nummer %in% list_cities_10)

cities_translation <- tibble(
  bfs_nummer = c(261, 6621, 2701, 5586, 351, 230, 1061, 3203, 5192, 371),
  name_en = c(
    "Zurich", "Geneva", "Basel", "Lausanne", "Bern",
    "Winterthur", "Lucerne", "St. Gallen", "Lugano", "Biel/Bienne"
  )
)

# Calculate city centroids
city_centroids <- sf_cities_10 %>%
  st_transform(4326) %>%  # WGS84 for Datawrapper
  st_centroid() %>%
  mutate(
    longitude = st_coordinates(.)[,1],
    latitude = st_coordinates(.)[,2]
  ) %>%
  st_drop_geometry() %>%
  select(bfs_nummer, name, longitude, latitude) %>%
  left_join(cities_translation, by = "bfs_nummer") %>%
  left_join(df_cities_pop, by = "bfs_nummer")

# Create mobile and desktop versions
dw_cities_situation_mobile <- city_centroids %>% 
  mutate(
    name = name_en,
    "text-fontSize" = 8,
    "text-color" = "#2c3e50",
    icon = "square",
    color = "#2c3e50",
    scale = scales::rescale(sqrt(POP_2023), to = c(0.5, 1.75)),
    mobile = TRUE,
    desktop = FALSE
  )

dw_cities_situation_desktop <- city_centroids %>% 
  mutate(
    name = name_en,
    "text-fontSize" = 12,
    "text-color" = "#2c3e50",
    icon = "square",
    color = "#2c3e50",
    scale = scales::rescale(sqrt(POP_2023), to = c(0.5, 2)),
    mobile = FALSE,
    desktop = TRUE
  )

dw_cities_situation <- rbind(dw_cities_situation_mobile, dw_cities_situation_desktop)

# Export to file (Datawrapper maps need manual CSV upload)
dir.create("data_output/dw_situation_map", showWarnings = FALSE, recursive = TRUE)
dw_cities_situation %>%
  write_csv("data_output/dw_situation_map/dw_cities_situation.csv")

cat("✅ Situation map data exported to data_output/dw_situation_map/dw_cities_situation.csv\n")
cat("ℹ️  Manual upload required to Datawrapper map 4bAiG\n")
```

# 5. Summary

```{r}
cat("\n📊 All Datawrapper charts updated:\n")
cat("- Qsh2p: National quartiles comparison\n")
cat("- BnMJJ: Cities aggregated bars\n") 
cat("- iLHkd: City disparities lollipop\n")
cat("- rffGE: St-Gallen NO₂ scatterplot\n")
cat("- 8ZPgH: Geneva NO₂ scatterplot\n")
cat("- qRJ8l: Geneva NDVI scatterplot\n")
cat("- MJu0Z: Lausanne noise scatterplot\n")
cat("- HfB0J: Lausanne NDVI scatterplot\n")
cat("- 4bAiG: Situation map (CSV exported for manual upload)\n")
cat("\n✅ All charts ready for publication!\n")
```
