---
title: "05_prepare_mapbox_utilities"
author: "Olaf König"
format: html
editor: visual
---

# Mapbox utilities and web assets preparation

## Objective

Prepare all auxiliary geographic data needed for the web map integration: anonymized SES hexagons, clean municipal boundaries, city centroids, and optionally classified raster tiles.

## Source data

-   `data_input/data_edit/cities_reclassified.gpkg` - Cities with SES classifications
-   `data_input/data_raw/sf_polg_swisstopo.gpkg` - Municipal boundaries
-   `data_input/data_raw/swisstopo_tlm_lakes.gpkg` - Lakes for boundary cleaning
-   `data_input/data_raw/stats_cities.csv` - City population data

## Processing steps

1.  **SES hexagon aggregation** : Create anonymized hexagonal grids showing dominant SES patterns
2.  **Municipal boundaries** : Clean city limits by removing lake areas
3.  **City centroids** : Calculate center points for map navigation and labeling
4.  **Situation map data** : Prepare responsive markers for overview map

## Final outputs

-   `data_output/mapbox/sep_hex.geojson` - Anonymized SES hexagons
-   `data_output/mapbox/cities.geojson` - Clean municipal boundaries
-   `data_output/city_centroids_mapbox.csv` - City center coordinates
-   `data_output/dw_situation_map/dw_cities_situation.csv` - Situation map markers

------------------------------------------------------------------------

```{r setup}
library(tidyverse)
library(sf)
```

```{r load-data}
# Load main datasets
sf <- read_sf("data_input/data_edit/cities_reclassified.gpkg") %>% 
  select(id, bfs_nummer, ssep3d_local_4)

sf_municipalities <- read_sf("data_input/data_raw/sf_polg_swisstopo.gpkg")
sf_lakes <- read_sf("data_input/data_raw/swisstopo_tlm_lakes.gpkg") %>% 
  filter(surface_km2 >= 1)

df_cities_pop <- read_csv("data_input/data_raw/stats_cities.csv") %>% 
  select(bfs_nummer, nom_officiel, POP_2023)

# Define city lists
list_cities_10 <- df_cities_pop %>% 
  slice_max(POP_2023, n = 10) %>% 
  pull(bfs_nummer)

list_cities_6 <- df_cities_pop %>% 
  slice_max(POP_2023, n = 6) %>% 
  pull(bfs_nummer)

sf_cities_10 <- sf_municipalities %>% 
  filter(bfs_nummer %in% list_cities_10)
```

# 1. SES hexagon aggregation for anonymization

## 1.1 Hexagon creation functions

```{r}
# Create hexagonal aggregation for one city
create_hex_city <- function(sf_city, hex_size = 100, min_points = 5) {
  
  city_code <- unique(sf_city$bfs_nummer)
  
  # Create hexagonal grid for this city
  hex_grid <- st_make_grid(
    sf_city, 
    cellsize = hex_size, 
    square = FALSE
  ) %>% 
    st_sf() %>% 
    mutate(hex_id = paste0(city_code, "_", row_number()))
  
  # Join points to hexagons
  sf_with_hex <- st_join(sf_city, hex_grid)
  
  # Calculate stats per hexagon
  hex_stats <- sf_with_hex %>% 
    st_drop_geometry() %>% 
    filter(!is.na(hex_id)) %>%
    group_by(hex_id, bfs_nummer) %>% 
    summarise(
      n_points = n(),
      pct_q1 = sum(ssep3d_local_4 == 1) / n() * 100,
      pct_q2 = sum(ssep3d_local_4 == 2) / n() * 100, 
      pct_q3 = sum(ssep3d_local_4 == 3) / n() * 100,
      pct_q4 = sum(ssep3d_local_4 == 4) / n() * 100,
      dominant_class = as.numeric(names(sort(table(ssep3d_local_4), decreasing = TRUE))[1]),
      dominant_pct = max(pct_q1, pct_q2, pct_q3, pct_q4),
      .groups = "drop"
    ) %>% 
    filter(n_points >= min_points) %>%
    mutate(
      dominance_type = case_when(
        dominant_pct >= 60 ~ paste0("Dominant Q", dominant_class),
        dominant_pct >= 40 ~ paste0("Mixed (Q", dominant_class, " majority)"),
        TRUE ~ "Highly mixed"
      )
    )
  
  # Final geometry with stats
  hex_final <- hex_grid %>% 
    inner_join(hex_stats, by = "hex_id")
  
  return(hex_final)
}

# Main aggregation function
create_hex_aggregation <- function(sf_data, hex_size = 100, min_points = 5) {
  
  # Split by city
  sf_by_city <- split(sf_data, sf_data$bfs_nummer)
  
  # Process each city
  hex_list <- map(sf_by_city, ~create_hex_city(.x, hex_size, min_points))
  
  # Combine results
  hex_combined <- bind_rows(hex_list)
  
  return(hex_combined)
}
```

## 1.2 Create and export SES hexagons

```{r}
cat("🔶 Creating anonymized SES hexagons...\n")

# Create hexagonal aggregation
sf_hex <- create_hex_aggregation(sf, hex_size = 100, min_points = 5)

# Summary statistics
cat("Hexagons created:", nrow(sf_hex), "\n")
sf_hex %>% 
  st_drop_geometry() %>% 
  count(dominance_type, sort = TRUE) %>%
  print()

# Export for web (top 3 cities only for demo)
sf_hex_mapbox <- sf_hex %>% 
  filter(bfs_nummer %in% c(3203, 5586, 6621)) %>%  # St. Gallen, Lausanne, Geneva
  select(bfs_nummer, n_points, dominant_class) %>% 
  st_transform(4326)

# Create output directory
dir.create("data_output/mapbox", showWarnings = FALSE, recursive = TRUE)

# Export
st_write(sf_hex_mapbox, "data_output/mapbox/sep_hex.geojson", delete_dsn = TRUE, quiet = TRUE)

cat("✅ SES hexagons exported to data_output/mapbox/sep_hex.geojson\n")
cat("   Cities included:", length(unique(sf_hex_mapbox$bfs_nummer)), "\n")
cat("   Total hexagons:", nrow(sf_hex_mapbox), "\n\n")
```

# 2. Municipal boundaries without lakes

## 2.1 Clean city boundaries

```{r}
cat("🗺️ Creating clean municipal boundaries...\n")

# Remove lakes from municipal boundaries
sf_cities_10_clean <- st_difference(sf_cities_10, st_union(sf_lakes))

# Transform to WGS84 and export
sf_cities_10_clean %>% 
  select(bfs_nummer, name) %>% 
  st_transform(4326) %>% 
  write_sf("data_output/mapbox/cities.geojson", delete_dsn = TRUE, quiet = TRUE)

cat("✅ Clean city boundaries exported to data_output/mapbox/cities.geojson\n")
cat("   Cities included:", nrow(sf_cities_10_clean), "\n")
cat("   Lakes removed from boundaries\n\n")
```

# 3. City centroids for map navigation

## 3.1 Calculate centroids and add metadata

```{r}
cat("📍 Calculating city centroids...\n")

# English translations
cities_translation <- tibble(
  bfs_nummer = c(261, 6621, 2701, 5586, 351, 230, 1061, 3203, 5192, 371),
  name_en = c(
    "Zurich", "Geneva", "Basel", "Lausanne", "Bern",
    "Winterthur", "Lucerne", "St. Gallen", "Lugano", "Biel/Bienne"
  )
)

# Calculate centroids (after lake removal)
city_centroids <- sf_cities_10_clean %>%
  st_transform(4326) %>%  # WGS84 for web
  st_centroid() %>%
  mutate(
    longitude = st_coordinates(.)[,1],
    latitude = st_coordinates(.)[,2]
  ) %>%
  st_drop_geometry() %>%
  select(bfs_nummer, name, longitude, latitude) %>%
  left_join(cities_translation, by = "bfs_nummer") %>%
  left_join(df_cities_pop, by = "bfs_nummer")

# Display results
print(city_centroids)

# Export for general use
write_csv(city_centroids, "data_output/city_centroids_mapbox.csv")

cat("✅ City centroids exported to data_output/city_centroids_mapbox.csv\n\n")
```

# 4. Situation map markers (Datawrapper format)

## 4.1 Create responsive markers for situation map

```{r}
cat("🗺️ Preparing situation map markers...\n")

# Mobile version (smaller text and markers)
dw_cities_situation_mobile <- city_centroids %>% 
  mutate(
    name = name_en,
    "text-fontSize" = 8,
    "text-color" = "#2c3e50",
    icon = "square",
    color = "#2c3e50",
    scale = scales::rescale(sqrt(POP_2023), to = c(0.5, 1.75)),
    mobile = TRUE,
    desktop = FALSE
  )

# Desktop version (larger text and markers)  
dw_cities_situation_desktop <- city_centroids %>% 
  mutate(
    name = name_en,
    "text-fontSize" = 12,
    "text-color" = "#2c3e50",
    icon = "square",
    color = "#2c3e50",
    scale = scales::rescale(sqrt(POP_2023), to = c(0.5, 2)),
    mobile = FALSE,
    desktop = TRUE
  )

# Combine responsive versions
dw_cities_situation <- rbind(dw_cities_situation_mobile, dw_cities_situation_desktop)

# Create output directory and export
dir.create("data_output/dw_situation_map", showWarnings = FALSE, recursive = TRUE)
write_csv(dw_cities_situation, "data_output/dw_situation_map/dw_cities_situation.csv")

cat("✅ Situation map markers exported to data_output/dw_situation_map/dw_cities_situation.csv\n")
cat("   Responsive versions: mobile + desktop\n")
cat("   Cities included:", length(unique(dw_cities_situation$bfs_nummer)) / 2, "\n\n")  # /2 because mobile+desktop
```

# 5. Optional: Raster tiles preparation

```{r eval=FALSE}
# This section is kept as reference but not executed by default
# Uncomment and modify if you need raster tiles for web map

## 5.1 Raster classification function
traiter_indicateur_par_ville <- function(raster, sf_villes, seuils, labels, nom_raster, sf_lakes) {
  cat("Processing", nom_raster, "\n")
  cat("  Raster CRS:", crs(raster, describe = TRUE)$name, "\n")
  
  # Process each city separately for manageable file sizes
  for (i in 1:nrow(sf_villes)) {
    sf_ville <- sf_villes[i, ]
    
    # Vectorize raster for this city
    sf_ville_vector <- vectoriser_raster_ville(
      raster = raster,
      sf_ville = sf_ville,
      seuils = seuils,
      labels = labels,
      nom_raster = nom_raster,
      sf_lakes = sf_lakes
    )
    
    # Save if data exists
    if (!is.null(sf_ville_vector) && nrow(sf_ville_vector) > 0) {
      ville_nom <- if("nom_officiel" %in% names(sf_ville)) sf_ville$nom_officiel else as.character(sf_ville$bfs_nummer)
      filename <- paste0("data_output/mapbox_layers/", nom_raster, "_", sf_ville$bfs_nummer, "_", 
                        gsub("[^A-Za-z0-9]", "", ville_nom), ".geojson")
      
      st_write(sf_ville_vector, filename, delete_dsn = TRUE, quiet = TRUE)
      
      size_kb <- round(file.size(filename) / 1024, 1)
      cat("    Saved:", basename(filename), "(", size_kb, "KB)\n")
    }
  }
}

# Example usage:
# rast_noise <- rast("data_input/data_raw/LrStrasseTag_4_0_Meter.tif")
# seuils_noise <- seq(0, 70, 10)
# labels_noise <- c("0-10", "10-20", "20-30", "30-40", "40-50", "50-60", "60-70", "70+")
# traiter_indicateur_par_ville(rast_noise, sf_cities_10, seuils_noise, labels_noise, "noise", sf_lakes)
```

# 6. Final summary

## 6.1 Mapbox assets summary

```{r}
cat("🎯 MAPBOX ASSETS SUMMARY\n")
cat("========================\n")

# Check all output files
mapbox_files <- list.files("data_output/mapbox", pattern = "\\.geojson$", full.names = TRUE)
csv_files <- list.files("data_output", pattern = "city_centroids_mapbox\\.csv", recursive = TRUE, full.names = TRUE)
situation_files <- list.files("data_output/dw_situation_map", pattern = "\\.csv$", full.names = TRUE)

all_files <- c(mapbox_files, csv_files, situation_files)

if (length(all_files) > 0) {
  total_size_kb <- 0
  
  for (file in all_files) {
    if (file.exists(file)) {
      size_kb <- round(file.size(file) / 1024, 1)
      total_size_kb <- total_size_kb + size_kb
      
      # Count features for geojson files
      if (grepl("\\.geojson$", file)) {
        n_features <- nrow(st_read(file, quiet = TRUE))
        cat("📁", basename(file), "-", size_kb, "KB -", n_features, "features\n")
      } else {
        cat("📁", basename(file), "-", size_kb, "KB\n")
      }
    }
  }
  
  cat("\nTotal size:", round(total_size_kb / 1024, 1), "MB\n")
} else {
  cat("❌ No output files found\n")
}

cat("\n🌍 All assets ready for web integration!\n")
```

## 6.2 Usage instructions

```{r}
cat("\n📖 USAGE INSTRUCTIONS\n")
cat("=====================\n")
cat("1. 🔶 sep_hex.geojson - Add as Mapbox layer for SES patterns\n")
cat("2. 🗺️ cities.geojson - Use as boundary overlay\n") 
cat("3. 📍 city_centroids_mapbox.csv - Use for map navigation/labeling\n")
cat("4. 🎯 dw_cities_situation.csv - Upload to Datawrapper situation map\n")
cat("\nAll files use WGS84 (EPSG:4326) coordinate system\n")
cat("Ready for web map integration! 🚀\n")
```
